`r opts_chunk$set(cache=TRUE, fig.width=8)`
<link href="http://joey711.github.com/phyloseq/markdown.css" rel="stylesheet"></link>

The phyloseq Demo Index
========================================================

This page is automagically updated when I do a periodic full rebuild of the phyloseq demo pages. The date of this particular re-build is `r date()`.

Don't forget to checkout [the phyloseq main page](joey711.github.com/phyloseq) for core tutorials.

---
## Available demos

```{r find-parse-demos, echo=FALSE, include=FALSE}
files = list.files(pattern=".[Rr]md")
length(files)
names(files) = unlist(strsplit(files, ".rmd", TRUE))
links = paste(names(files), ".html", sep="")
mdlines = paste("### [", names(files), "](", links, ")", sep="")
```

```{r write-links, results='asis', echo=FALSE}
cat(mdlines, sep="\n\n")
```


---
## The R Markdown Source Code
For more advanced users interested in seeing and/or modifying the [R markdown](http://www.rstudio.com/ide/docs/r_markdown) source code from which these reproducible, web browser-friendly demos were created.

```{r rmd-links, results='asis', echo=FALSE}
rmdlines = paste("### [", names(files), "](", files, ")", sep="")
cat(rmdlines, sep="\n\n")
```