 Demo: phyloseq – A Bioconductor package for handling and analysis of high-throughput phylogenetic sequence data 
====================================================================================================================

 Paul J. McMurdie and Susan Holmes
    Statistics Department, Stanford University,
    Stanford, CA 94305, USA
    
    E-mail: mcmurdie@stanford.edu
    https://github.com/joey711/phyloseq
    susan@stat.stanford.edu\
    http://www-stat.stanford.edu/~susan/

Summary
---------

This is a demonstration manual for the phyloseq package. It is a Sweave
source-file with example code-chunks that can be reused if you have
phyloseq installed.

phyloseq Documentation
-------------------

Vignettes are included in phyloseq. A quick way to load them from within `R` is:

```{r, eval=FALSE}
vignette("phyloseqbasics")
vignette("phyloseqanalysis")
```

This demonstration document supports a live demonstration of tools in
phyloseq, and supplements other documentation resources
available for [the phyloseq package](https://github.com/joey711/phyloseq) (e.g. wiki, vignettes, publications, function-level documentation, etc.).

There is also package-level documentation available. Note the following difference:
```{r, eval=FALSE}
?"phyloseq-package"
?phyloseq # this is a function
```

Importing Data
-------------------


